Live Functional Programming with Typed Holes

### 2018-04-08 calculus.tex
comments with small suggestions in Section 3

For the familiar forms of the lambda calculus, the rules again follow prior work. For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e. $b$ is the unit type). \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}.

Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation. The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}. We will return to evaluation in Sec.~\ref{sec:evaluation}.

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. We say more about hole contexts below. The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system: 
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} (so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation onto unannotated external lambdas based on the given type. The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.

\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively.

@@

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$. The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. Initially, no evaluation has occured, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed. The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion.

@@

Casts are inserted during the expansion of function applications and ascriptions. The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.

@@

The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}. The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}

@@

Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. The only cases of special interest are when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write $[d/x]\sigma$ to perform substitution on each expression in the hole environment, $\sigma$. For example, $\stepsToD{}{\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}{\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g. because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?} Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.

>>>>

For the familiar forms of the lambda calculus, the rules again follow prior work. For simplicity, the core calculus includes only a single base type, $b$, with a single constant, $c$, governed by Rule {SConst} (i.e. $b$ is the unit type). \Hazelnut instead defined a number type with a single operation, which we include in Appendix~\ref{sec:extensions} alongside various other standard extensions to the core calculus\todo{do this, say more?}. 
%
\rkc{perhaps too much info / unnecessary to say Hazelnut had numbers instead of unit?} 
%
Rule {SVar} specifies that variables synthesize the corresponding type from $\hGamma$.

@@

Well-typed external expressions expand to well-typed internal expressions, $d$, for evaluation. The rules governing expansion are given in Fig.~\ref{fig:expansion} and the rules governing type assignment for internal expressions are given in Fig.~\ref{fig:hasType}. Evaluation of internal expressions will be discussed in Sec.~\ref{sec:evaluation}.

Expansion is specified, much like the type system for the external language, in the bidirectional style by two mutually defined judgements. The synthetic expansion judgement, $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$, synthesizes a type, $\htau$, from $\hexp$, and produces an expansion, $d$, and a hole context, $\hDelta$. We say more about hole contexts below. The analytic expansion judgement, $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks $\hexp$ against $\htau$ and produces an expansion, $d$, of type $\htau'$, and a hole context, $\hDelta$. The governing theorem below establishes that the type $\htau'$ is necessarily consistent with $\htau$.
%
\rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
entirely standard, because of the hole context, so perhaps worth mentioning.}
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent
The reason analytic expansion produces an expansion of consistent type is because the subsumption rule, as previously discussed, allows us to check an external expression against any type consistent with the type the expression actually synthesizes, whereas every internal expression can be assigned at most one type, i.e. the following standard unicity property holds of the type assignment system: 
\begin{thm}[Type Assignment Unicity]
  If $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$
  and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$
  then $\htau=\htau'$.
\end{thm}
\noindent
Consequently, analytic expansion reports the type actually assigned to the expansion it produces. For example, we can derive that $\expandAna{\hGamma}{c}{\tehole}{c}{b}{\emptyset}$.% where $\emptyset$ is the empty hole context.

@@

The rules governing expansion of constants, variables and lambda expressions --- Rules {ESConst}, {ESVar}, {ESLam} and {EALam} --- and the corresponding type assignment rules --- Rules {TAConst}, {TAVar} and {TALam} --- mirror the typing rules from Fig.~\ref{fig:bidirectional-typing} (so the corresponding cases of Theorem~\ref{thm:typed-expansion}, Theorem~\ref{thm:expandability} and Theorem~\ref{thm:expansion-generality} are straightforward). Note that in the internal language, all lambdas are half-annotated, again to support type assignment---Rule {EALam} inserts the annotation onto unannotated external lambdas based on the given type. The rules governing expansion of holes, function application and ascription are more interesting, so let us consider them in turn.
%
\rkc{this list (holes, func app, and ascription) has three items, but there are
only two subsections. see comment at beginning of Sec 3.2.2.}

\subsubsection{Hole Expansion}\label{sec:hole-expansion} Rules {ESEHole}, {ESNEHole}, {EAEHole} and {EANEHole} govern the expansion of empty and non-empty expression holes to \emph{hole closures}, $\dehole{u}{\sigma}{}$ and $\dhole{\dexp}{u}{\sigma}{}$, respectively. 
%
\rkc{perhaps add ``empty and non-empty'' before hole closures.}

@@

Each hole closure also has an associated environment, $\sigma$, which is a finite substitution of the form $[d_1/x_1, ~\cdots, d_n/x_n]$ for $n \geq 0$.  The purpose of the closure environment is to keep a record of the substitutions that occur around the hole as evaluation occurs. Initially, no evaluation has occurred, so the initial environment generated by the hole expansion rules is the identity substitution for the typing context associated with $u$ in $\Delta$, which we notate $\idof{\hGamma}$ and define as follows.
\begin{defn}[Identity Substitution] $\idof{x_1 : \tau_1, ~\cdots, x_n : \tau_n} = [x_1/x_1, ~\cdots, x_n/x_n]$
\end{defn}
\noindent
The type assignment rules for hole closures, Rules TAEHole and TANEHole, require that we be able to check the environment of each hole closure against the corresponding typing context, written $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ and defined as follows:\todo{check that the definition in the Agda corresponds}
\begin{defn}[Substitution Typing]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff $\domof{\sigma} = \domof{\hGamma'}$ and for each $x : \htau \in \hGamma'$ we have that $d/x \in \sigma$ and $\hasType{\Delta}{\hGamma}{d}{\tau}$.
\end{defn}
\noindent
It is easy to verify that the identity substitution satisfies this requirement, i.e. that $\hasType{\Delta}{\hGamma}{\idof{\hGamma}}{\hGamma}$.

@@

\subsubsection{Cast Insertion}\label{sec:cast-insertion} 
%
\rkc{consider renaming subsection to ``Function Application and Cast
Intersion''. and adding a transition sentence like ``The remaining development
of expansion deals with the delicate interplay between function application and
cast insertion.''}
%
Consider the following example: $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$. The type synthesized for this example viewed as an external expression is $\tehole$, because the hole type annotation on $x$ allows us to apply it as a function of type $\tarr{\tehole}{\tehole}$, as previously discussed, and $c$ can be checked against type $\tehole$ by subsumption. However, viewed as an internal expression, this example is not well-typed---we do not have subsumption in the type assignment system defined in Fig.~\ref{fig:hasType}. Indeed, it would violate type safety if we could assign a type to this example in the internal language, because beta reduction of this example viewed as an internal expression would result in $c(c)$, which is clearly not well-typed. The difficulty is that leaving the argument type unknown leaves how the argument is being used (in this case, as a function) also unknown.\footnote{In a system where type reconstruction is first used to try to fill in type holes, we could express a similar example by using $x$ at two or more different types, thereby causing type reconstruction to fail.
%On the other hand, if it is acceptable to arbitrarily choose one of the possible types, and type reconstruction is complete, then type holes will never appear in the internal language and the cast insertion machinery described in this section can be omitted entirely, leaving only the hole closure machinery described previously.
}\todo{talk about elsewhere? maybe do type-hole-free version of calculus in appendix if there is time}{} Recall that we can interpret hole types as unknown types from gradual type theory. This leads us to the solution to this problem: cast insertion. 
%
\rkc{Instead of the last two sentences, maybe something like: ``By analogy
between hole types in our system and unknown types from gradual type theory, we
address the problem with the same mechanism used in gradual type theory: cast
insertion.''}

@@

Casts are inserted during the expansion of function applications and ascriptions. The latter is more straightforward: Rule~{ESAsc} in Fig.~\ref{fig:expandSyn} inserts a cast from the assigned type to the ascribed  type. Theorem~\ref{thm:typed-expansion} inductively ensures that the two types are consistent.  Note that we included ascription only for the sake of exposition: it can be defined using application together with the half-annotated identity function: $e : \tau = \hap{(\halam{x}{\htau}{x})}{e}$.
%
\rkc{move this Note sentence to the same place where half-annotated lambdas were
noted. This note breaks the flow here.}

@@

The judgement $\isIndet{d}$ distinguishes \emph{indeterminate} forms, so named because they arise from the presence of expression holes and failed casts in a program, and so, conceptually, their ultimate value awaits programmer action (see Sec.~\ref{sec:commutativity}). The first two rules specify that hole closures are indeterminate (in the case of non-empty hole closures, when the inner expression is final). The rules defining the remaining indeterminate forms are explained simultaneously with the corresponding transition rules below.\todo{relationship to weak head normal forms}

@@

Substitution, $[d/x]d'$, operates in the standard capture-avoiding manner \cite{pfpl}. The only cases of special interest are when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write $[d/x]\sigma$ to perform substitution on each expression in the hole environment, $\sigma$. For example, $\stepsToD{}{\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}{\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$. Hole closures can be duplicated like any other term by substitution, and the environments of different closures for the same hole name can differ, e.g. because a function with a hole closure in its body is applied multiple times.\todo{show example here or refer back to sec 2?}. Hole closures can also appear within the environments of other hole closures (this gives rise to the closure paths described in Sec.~\ref{sec:paths}).% If we uniquely number each hole closure in an expression, then each unique hole closure can be identified by a path consisting of a hole name, an instance number and a variable.
